---
title: "Assignment 8: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Tristan Frappier-Brinton"
date: "Spring 2026"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**

>NOTE: For this assignment, you can skip tests for any of the assumptions required to run the models. 

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae)   
library(here)
here()

#Read in the processed EPA air quality dataset
LTER <- read.csv(
  here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
  stringsAsFactors = TRUE)


#2
my_theme <- 
  theme(
    #line =              element_line(),
    #rect =              element_rect(),
    text =              element_text(size=10),
    
    # Modified inheritance structure of text element
    #plot.title =        element_text(),
    axis.title.x =      element_text(color="lightgrey"),
    axis.title.y =      element_text(color="lightgrey"),
    axis.text =         element_text(color="lightgrey"),
    
    # Modified inheritance structure of line element
    axis.ticks =        element_line(color = 'lightgrey'),
    #panel.grid.major =  element_line(),
    #panel.grid.minor =  element_blank(),
    
    # Modified inheritance structure of rect element
    plot.background =   element_rect( color='black',
      fill = 'black'),
    legend.background = element_rect(
      color='orange',
      fill = 'darkgray'),
    #panel.background =  element_rect(),
    #legend.key =        element_rect(),
    
    # Modifiying legend.position
    #legend.position = 'right',
    
    #complete = TRUE
    )

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Lake temperature in July does not change with depth across all lakes.
Ha: Lake temperature in July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
LTER2 <- LTER %>% 
  select(`lakename`, `year4`, `daynum`, `depth`, `temperature_C`) %>%
  filter(daynum >181, daynum <213) %>%
  drop_na()

#5

plot1 <- ggplot(LTER2, aes(x = depth, y = temperature_C)) + 
  geom_point(alpha=0.5) + 
  geom_smooth(method = "lm") +
  my_theme +
  ylim(0, 35) +  
  ylab("Temperature (C)") +
  xlab("Depth (m)")
plot1
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Temperature is highly variable at low depths, likely more closely resembling the temperature of the air (15-30 degrees C). Both temperature and variability become lower at lower depths, until temperatures start to plateau below 10 m at roughly 5 degrees C. Thus, this is a non-linear trend.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
lm1 <- lm(data=LTER2, temperature_C ~ depth)
summary(lm1)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
There is a significant relationship between temperature and depth. Specifically, the model estimates that temperature becomes significantly lower (by 1.94 *C) for every increase of 1m in depth. 73.91% of the variability in temperature is explained by depth. This finding is based on 9720 degrees of freedom. The relationship between temperature and depth is highly significant, with the model estimating there is a less that 0.00000000000002% chance that such a relationship would emerge from chance alone in independent data. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
lm2 <- lm(data=LTER2, temperature_C ~ depth + daynum + year4)
step(lm2) #lowest AIC is when no variables are removed


#10
lm2 <- lm(data=LTER2, temperature_C ~ depth + daynum + year4)
summary(lm2)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC values suggest that we use all of year4, daynum, and depth as explanatory variables in our regression. This resulting regression explains 74.17% of the variation in temperature. This is a slight improvement over the model only using depth.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
anova1 <- aov(data = LTER2, temperature_C ~ lakename)
summary(anova1)

lm3 <- lm(data=LTER2, temperature_C ~ lakename)
summary(lm3)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a highly significant difference in mean temperature among the lakes. This is based on an anova with 8, 9713 degrees of freedom, and p value below 2e-16.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
plot2 <- ggplot(LTER2, aes(x = depth, y = temperature_C, color=lakename), group_by(lakename)) + 
  geom_point(alpha=0.5) + 
  geom_smooth(method = "lm", se=FALSE) +
  my_theme +
  ylim(0, 35) +  
  ylab("Temperature (C)") +
  xlab("Depth (m)") +
  labs(color="Lake Name")
plot2

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(anova1)

HSDgroups <- HSD.test(anova1, "lakename", group=TRUE)
HSDgroups
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul and Ward Lakes have mean temperatures that are not significantly different from that of Peter Lake. No lake has a mean temperature that is significantly different from all other lakes. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: two-sample t-test



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
LTER3 <- LTER2 %>%
  filter(lakename=="Crampton Lake" | lakename == "Ward Lake")

ttest <- lm(data=LTER3, temperature_C ~ lakename)
summary(ttest)
```

>Answer: The resulting p value of 0.259 suggests that the mean temperatures are not significantly different between the two lakes, and this does match the results above, where both lakes were included in group b.
