---
title: "Assignment 10: Data Scraping"
author: "Tristan Frappier-Brinton"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)

getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
url <- read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024")

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
water_system <- url %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()


contact_person <- url %>% 
  html_nodes("div+ table tr:nth-child(4) td:nth-child(2)") %>%
  html_text()

ownership <- url %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()

mgd <- url %>% html_nodes("th~ td+ td") %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
dw <- data.frame(WaterSystem = water_system,
                 ContactPerson = contact_person,
                 Ownership = ownership,
                 MaximumDailyUse = mgd,
                 Month=c(1,5,9,2,6,10,3,7,11,4,8,12),
                 Year=rep("2024"))

dw$Date <- ymd(paste(dw$Year, dw$Month, "1", sep = "-"))

dw$Month <- factor(dw$Month,
                   levels = 1:12,
                   labels = month.abb)

#5 

ggplot(dw,aes(x=as.factor(Month),y=as.numeric(MaximumDailyUse), group=1)) +
  geom_line( color="red")+
  xlab("Calendar Month")+
  ylab("Maximum Daily Water Use (MGD)")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
scrape.it <- function(year, PWSID){

url <- read_html(paste0(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", 
  PWSID, "&year=",year))
                 
water_system <- url %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()

contact_person <- url %>% 
  html_nodes("div+ table tr:nth-child(4) td:nth-child(2)") %>%
  html_text()

ownership <- url %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()

mgd <- url %>% html_nodes("th~ td+ td") %>% html_text()



dw <- data.frame(WaterSystem = water_system,
                 ContactPerson = contact_person,
                 Ownership = ownership,
                 MaximumDailyUse = mgd,
                 Month=c(1,5,9,2,6,10,3,7,11,4,8,12),
                 Year=year,
                 PWSID = PWSID)

dw$Date <- ymd(paste(dw$Year, dw$Month, "1", sep = "-"))

dw$Month <- factor(dw$Month,
                   levels = 1:12,
                   labels = month.abb)

return( dw)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
df_scraped <- scrape.it(year="2020", PWSID="03-32-010")

```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
df_Cary <- scrape.it(year="2020", PWSID="03-92-020")
df_combined <- rbind(df_scraped, df_Cary)

ggplot(df_combined,aes(
  x=as.factor(Month),y=as.numeric(MaximumDailyUse), group = WaterSystem)) +
  geom_line( aes(color=WaterSystem))+
  xlab("Calendar Month")+
  ylab("Maximum Daily Water Use (MGD)")
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 

#df_range <- map(c(2018:2023),scrape.it,PWSID="03-92-020") %>% bind_rows()
df_range <- map2(c(2018:2023),"03-92-020",scrape.it) %>% bind_rows()

df_range$Month <- factor(df_range$Month,
                   labels = 1:12,
                   levels = month.abb)

df_range$Date <- ymd(paste(df_range$Year, df_range$Month, "1", sep = "-"))

ggplot(df_range,aes(x=Date,y=as.numeric(MaximumDailyUse), group = 1)) +
  geom_line( )+
  geom_smooth(method=loess)
  xlab("Date") +
  ylab("Maximum Daily Water Use (MGD)")


```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: Yes, Cary seems to have a slight trend (at this time scale) for increasing water usage.
>
