---
title: "Assignment 3: Data Exploration"
author: "Tristan Frappier-Brinton"
date: "Spring 2026"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  [NEW] Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to Canvas.

8.  Initial here to acknowledge that you did not use AI in completing this assignment, except where expressly allowed: TFB

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks in your code chunks.

**TIP**: If your code fails to knit, check:
 * That no `install.packages()` or `View()` commands exist in your code. 
 * That you are not displaying the entire contents of a large dataframe in your code.

---

## Set up your R session

1.  Load necessary packages (tidyverse, here), check your current working directory and import two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 

**Be sure to**:
* Use the `here()` package in specifying the paths to your datasets
* Include the appropriate subcommand to read in character based columns as factors


```{r, Import data, message = FALSE}
library(tidyverse)
library(here)

here()
Neonics <- read.csv(
  file = here("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"),
  stringsAsFactors = TRUE)   

Litter <- read.csv(
  file = here("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
  stringsAsFactors = TRUE)  
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. (AI is allowed here, but put answers in your own words.)

> Answer: Because insects are important pollinators and food chain members and we are/were worried that neonicotinoids were killing them.


3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. (AI is allowed here, but put answers in your own words.)

> Answer: Many insects make their homes in leaf litter. Litter could be sampled as a means of sampling the insect population, as a way of controlling for other factors (like litter depth) that may be influencing insects separately from neonicotinoids, or the litter itself could be tested for chemicals.


4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter and woody debris are collected from elevated and ground traps. Plant parts are sorted by type and weeighed to within 0.01g
 2. Each elevated trap is a 0.5m^2 square with mesh ‘basket’ elevated
~80cm above the ground. 
 3. Ground traps are 3 m x 0.5 m rectangular areas.


## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset? 4623 rows by 30 columns

```{r, dimensions}
summary(Neonics)
dim(Neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied.  [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r, Most common effects}
sort(summary(Neonics$Effect))
#the most common effect is "Population", then "Mortality"
```

> Question: Which two effects stand out as the most studied? Can you guess why these effects might specifically be of interest?
> Answer:"Population", then "Mortality". These might be of interest as there are worries/a priori hypotheses that neonicotinoids may increase insect mortality and decrease population size, which would be harmful to ecosystems and agriculture.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name).[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r, most studied species}
colnames(Neonics)
sort(summary(Neonics$Species.Common.Name))
#6 most common species are: Honey Bee, Parasitic Wasp, Buff tailed bumblebee, Carniolan honey bee, bumble bee, and intalian honeybee.
```

> Question: What do these species have in common? Why might they be of interest over other insects? 
> Answer: These are all hymenopteran pollinators, of particular interest because agriculture relies on pollinators


8.  The `Conc.1..Author` column, which lists the concentration of the neonicitoid dose, should include numeric values. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r, concentration}
class(Neonics$Conc.1..Author.)

```

> Answer: It is not numeric, it is a factor. This is because many entries have non-numerical symbols in them like "~" and ">".


## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r, number of studies freqpoly}

ggplot(Neonics, aes(Publication.Year)) +
  geom_freqpoly(binwidth = 1)
```


10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r, Test Location}
ggplot(Neonics, aes(Publication.Year, color=Test.Location)) +
  geom_freqpoly(binwidth = 1)
```


>Interpret this graph. What are the most common test locations, and do they differ over time?
> Answer: Lab and Field natural are the most common test locations. Field natural was briefly thee most common ~ 2009, but Lab studies have been the most common by far since 2010.


11. Create a bar graph of Endpoint counts. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r, endpoints}
ggplot(Neonics, aes(x=Endpoint)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#fails to eexecut ewith the message "Error in list2(na.rm = na.rm, binwidth = binwidth, bins = bins, orientation = orientation,  : object 'ffi_list2' not found"



```


> What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix (p.721) for more information. 
> Answer:NOEL and LOEL. NOEL is when a high concentration has no detectable effect compared to controls, and LOEL is the lowest concentration producing significant effeects compared to the controls.

---

## Explore your data (Litter)

12. Determine the class of `collectDate`. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r, Litter}
class(Litter$collectDate)#it is a factor
unique(Litter$collectDate) #Aug 2 and Aug 30 of 2018

Litter$collectDate <- ymd(Litter$collectDate)
class(Litter$collectDate)#it is a Date now
```


13. Using the `unique` function, list the different `plotIDs` sampled at Niwot Ridge. 

```{r, plotIDs}
unique(Litter$plotID)
summary(Litter$plotID)

```

> How is the information obtained from `unique` different from that obtained from `summary`?
> Answer:Unique just lists the possiblee names of plotID, summary also lists the frequency of each name


14. Create a bar graph of `functionalGroup` counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r, functionalGroups}
ggplot(Litter, aes(x=functionalGroup)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



```


15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of `dryMass` by `functionalGroup`.

```{r, drymass x functionalGroup}
ggplot(Litter, aes(x=functionalGroup,y=dryMass)) +
  geom_boxplot() 

ggplot(Litter, aes(x=functionalGroup,y=dryMass)) +
  geom_violin() 

```

> Why is the boxplot a more effective visualization option than the violin plot in this case?
> Answer: Overall point density is incredibly low relative to the maximum/spread of the data.


> What type(s) of litter tend to have the highest biomass at these sites?
> Answer:Needles tend to have the highest biomass.
