---
title: "Assignment 2: Coding Basics"
author: "Tristan Frappier-Brinton"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document. 
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

7.  Initial here to acknowledge that you did not use AI at all in completing this assignment: TFB


## Basics, Part 1

1.  Use R's `seq()` function to create a sequence of numbers from 100 to 333, increasing by threes. Assign this sequence a variable name.

2.  Compute the *mean* of this sequence, assigning this values its own variable name. 

3.  Compute the *standard deviation* (`sd()`) of this sequence, assigning this values its own variable name.  

4.  Display the the mean minus the standard deviation as well as the mean plus the standard deviation.

5.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. create a sequence from 100-333, counting by 3
a <- seq(from=100, to =333,by = 3)

#2. calculate mean of sequence
mean_a <- mean(a)

#3. calculate standard deviation of sequence
sd_a <- sd(a)

#4. show thee mean + and - thee standard deviation
mean_a - sd_a
mean_a + sd_a
```

---

## Basics, Part 2
6.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

7.  Label each vector with a comment on what type of vector it is.

8.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

9.  Label the columns of your data frame with informative titles.

```{r basics 2}

StudentName <- c("Bob", "Bobby", "Robert", "Bobert")
TestScores <- c(100, 75, 80, 55)
Scholarship <- c(TRUE, FALSE, TRUE, FALSE)

student_results <- cbind(StudentName, TestScores, Scholarship) 
student_results <- as.data.frame(student_results)
```

10.  QUESTION: How is this data frame different from a matrix?

> Answer: the columnc/vectors each have diffeerent data types

---

## Basics, Part 3

11. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, it returns the word "Pass"; otherwise it returns the word "Fail". 

12. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `. 

13. Run both functions using the value 54 as the input

14. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

```{r basics 3}
#11. Create a function using if...else
q3.1 <- function(x){
  if (x>50) {"Pass"
} else { "Fail"}
}

#12. Create a function using ifelse()
q3.2 <- function(x){ ifelse(x>50,"Pass","Fail")
}

#13a. Run the first function with the value 54
q3.1(54)
#13b. Run the second function with the value 54
q3.2(54)
#14a. Run the first function with the vector of test scores
#q3.1(student_results$TestScores)

#14b. Run the second function with the vector of test scores
q3.2(student_results$TestScores)
```

15. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization"; it's ok here if an AI response is presented in the search response.)

> Answer: only the ifelse() statement worked beecause it is designed to work with vectors while if{} and else{} are not


**NOTE** Before knitting, you'll need to comment out the call to the function in Q14 that does not work. (A document can't knit if the code it contains causes an error!)